---
title: "Homework 8"
author: "Xinyi Lin"
date: "11/17/2019"
output:
  word_document: default
  pdf_document: default
---

```{r,message=FALSE}
library(tidyverse)
library(survey)
```

```{r}
hw8_data = read.csv("./data_hw8.csv")
head(hw8_data)
```

## Question 1

1. Time point 1: L1, L2, L3, A

2. Time point 2: L1, L2, A

3. Time point 3: L1, L2, A

4. Time point 4: L1, L2, A

5. Time point 5: L1, L2, A

6. Time point 6: L1, L2, A, Y

## Question 2

The causal contrast is:

$E[Y|A_i=1]-E[Y|A_i=0]$, for $i=0,1,2,3,4,5,6$.

## Question 3

Let $L_{ij}$ indicates covariate $L_i$ in time point $j$.

## Question 4

(i) No unmeasured exposure-outcome confounding given C

(ii) No unmeasured mediator-outcome confounding given C

(iii) No unmeasured exposure-mediator confounding given C

(iv) No effect of exposure that confounds the mediator-outcome relationship

## Question 5

```{r}
# create wide data
AC_data = hw8_data %>% 
  gather(key = "Lt", value = "value", c(L1,L2,A)) %>% 
  arrange(id) %>% 
  mutate(Lt = str_c(Lt, t0)) %>% 
  select(id, Lt, value) %>% 
  spread(key = Lt, value = value)
Y_data = hw8_data %>% 
  select(id, L3, Y) %>% 
  na.omit()
wide_data = merge(AC_data, Y_data)
```

```{r}
# Time point 0
glm.model0 = glm(A0~L3, data = wide_data, family = binomial)
p0 = predict(glm.model0, type = "response")
w0 = ifelse(wide_data$A0==1, 1/p0, 1/(1-p0))
# Time point 1
glm.model1 = glm(A1~L3+A0+L10+L20, data = wide_data, family = binomial)
p1 = predict(glm.model1, type = "response")
w1 = ifelse(wide_data$A1==1, 1/p1, 1/(1-p1))
# Time point 2
glm.model2 = glm(A2~A1+L11+L21+L3+A0+L10+L20, data = wide_data, family = binomial)
p2 = predict(glm.model2, type = "response")
w2 = ifelse(wide_data$A2==1, 1/p2, 1/(1-p2))
# Time point 3
glm.model3 = glm(A3~A2+L12+L22+A1+L11+L21+L3+A0+L10+L20, data = wide_data, family = binomial)
p3 = predict(glm.model3, type = "response")
w3 = ifelse(wide_data$A3==1, 1/p3, 1/(1-p3))
# Time point 4
glm.model4 = glm(A4~A3+L13+L23+A2+L12+L22+A1+L11+L21+L3+A0+L10+L20, data = wide_data, family = binomial)
p4 = predict(glm.model4, type = "response")
w4 = ifelse(wide_data$A4==1, 1/p4, 1/(1-p4))
# Time point 5
glm.model5 = glm(A5~A4+L14+L24+A3+L13+L23+A2+L12+L22+A1+L11+L21+L3+A0+L10+L20, data = wide_data, family = binomial)
p5 = predict(glm.model5, type = "response")
w5 = ifelse(wide_data$A5==1, 1/p5, 1/(1-p5))
# Time point 6
glm.model6 = glm(A6~A5+L15+L25+A4+L14+L24+A3+L13+L23+A2+L12+L22+A1+L11+L21+L3+A0+L10+L20, data = wide_data, family = binomial)
p6 = predict(glm.model6, type = "response")
w6 = ifelse(wide_data$A6==1, 1/p6, 1/(1-p6))
w = w0*w1*w2*w3*w4*w5*w6
```

# Question 6

Marginal model: $E[Y_a]=\beta+\beta_0A_0+\beta_1A_1+\beta_2A_2+\beta_3A_3+\beta_4A_4+\beta_5A_5+\beta_6A_6$

```{r}
set.seed(123)
nboots = 1000
n_sample = nrow(wide_data)

beta = rep(NA, nboots)
beta0 = rep(NA, nboots)
beta1 = rep(NA, nboots)
beta2 = rep(NA, nboots)
beta3 = rep(NA, nboots)
beta4 = rep(NA, nboots)
beta5 = rep(NA, nboots)
beta6 = rep(NA, nboots)

for (i in 1:nboots) {
  S.b <- sample(1:n_sample, size = n_sample, replace = TRUE)
  data.b <- wide_data[S.b, ]
  # Time point 0
  glm.model0 = glm(A0~L3, data = data.b, family = binomial)
  p0 = predict(glm.model0, type = "response")
  w0 = ifelse(data.b$A0==1, 1/p0, 1/(1-p0))
  # Time point 1
  glm.model1 = glm(A1~L3+A0+L10+L20, data = data.b, family = binomial)
  p1 = predict(glm.model1, type = "response")
  w1 = ifelse(data.b$A1==1, 1/p1, 1/(1-p1))
  # Time point 2
  glm.model2 = glm(A2~A1+L11+L21+L3+A0+L10+L20, data = data.b, family = binomial)
  p2 = predict(glm.model2, type = "response")
  w2 = ifelse(data.b$A2==1, 1/p2, 1/(1-p2))
  # Time point 3
  glm.model3 = glm(A3~A2+L12+L22+A1+L11+L21+L3+A0+L10+L20, data = data.b, family = binomial)
  p3 = predict(glm.model3, type = "response")
  w3 = ifelse(data.b$A3==1, 1/p3, 1/(1-p3))
  # Time point 4
  glm.model4 = glm(A4~A3+L13+L23+A2+L12+L22+A1+L11+L21+L3+A0+L10+L20, data = data.b, family = binomial)
  p4 = predict(glm.model4, type = "response")
  w4 = ifelse(data.b$A4==1, 1/p4, 1/(1-p4))
  # Time point 5
  glm.model5 = glm(A5~A4+L14+L24+A3+L13+L23+A2+L12+L22+A1+L11+L21+L3+A0+L10+L20, data = data.b, family = binomial)
  p5 = predict(glm.model5, type = "response")
  w5 = ifelse(data.b$A5==1, 1/p5, 1/(1-p5))
  # Time point 6
  glm.model6 = glm(A6~A5+L15+L25+A4+L14+L24+A3+L13+L23+A2+L12+L22+A1+L11+L21+L3+A0+L10+L20, data = data.b, family = binomial)
  p6 = predict(glm.model6, type = "response")
  w6 = ifelse(data.b$A6==1, 1/p6, 1/(1-p6))
  w = w0*w1*w2*w3*w4*w5*w6
  
  data.b$w = w
  design = svydesign(ids = ~id, weights = ~w, data = data.b)
  msm = svyglm(Y ~ A0 + A1 + A2 + A3 + A4 + A5 + A6, family = gaussian(link = "identity"), design = design)
  beta[i] = msm$coef[1]
  beta0[i] = msm$coef[2]
  beta1[i] = msm$coef[3]
  beta2[i] = msm$coef[4]
  beta3[i] = msm$coef[5]
  beta4[i] = msm$coef[6]
  beta5[i] = msm$coef[7]
  beta6[i] = msm$coef[8]
}
```

```{r}
beta_est = mean(beta)
CIL_beta = mean(beta) - 1.96*sqrt(var(beta))
CIU_beta = mean(beta) + 1.96*sqrt(var(beta))
beta0_est = mean(beta0)
CIL_beta0 = mean(beta0) - 1.96*sqrt(var(beta0))
CIU_beta0 = mean(beta0) + 1.96*sqrt(var(beta0))
beta1_est = mean(beta1)
CIL_beta1 = mean(beta1) - 1.96*sqrt(var(beta1))
CIU_beta1 = mean(beta1) + 1.96*sqrt(var(beta1))
beta2_est = mean(beta2)
CIL_beta2 = mean(beta2) - 1.96*sqrt(var(beta2))
CIU_beta2 = mean(beta2) + 1.96*sqrt(var(beta2))
beta3_est = mean(beta3)
CIL_beta3 = mean(beta3) - 1.96*sqrt(var(beta3))
CIU_beta3 = mean(beta3) + 1.96*sqrt(var(beta3))
beta4_est = mean(beta4)
CIL_beta4 = mean(beta4) - 1.96*sqrt(var(beta4))
CIU_beta4 = mean(beta4) + 1.96*sqrt(var(beta4))
beta5_est = mean(beta5)
CIL_beta5 = mean(beta5) - 1.96*sqrt(var(beta5))
CIU_beta5 = mean(beta5) + 1.96*sqrt(var(beta5))
beta6_est = mean(beta6)
CIL_beta6 = mean(beta6) - 1.96*sqrt(var(beta6))
CIU_beta6 = mean(beta6) + 1.96*sqrt(var(beta6))
```

The estimated value of $\beta$ is `r round(beta_est,3)` and 95% confidence interval is (`r round(CIL_beta,3)`, `r round(CIU_beta,3)`).

The estimated value of $\beta0$ is `r round(beta0_est,3)` and 95% confidence interval is (`r round(CIL_beta0,3)`, `r round(CIU_beta0,3)`).

The estimated value of $\beta1$ is `r round(beta1_est,3)` and 95% confidence interval is (`r round(CIL_beta1,3)`, `r round(CIU_beta1,3)`).

The estimated value of $\beta2$ is `r round(beta2_est,3)` and 95% confidence interval is (`r round(CIL_beta2,3)`, `r round(CIU_beta2,3)`).

The estimated value of $\beta3$ is `r round(beta3_est,3)` and 95% confidence interval is (`r round(CIL_beta3,3)`, `r round(CIU_beta3,3)`).

The estimated value of $\beta4$ is `r round(beta4_est,3)` and 95% confidence interval is (`r round(CIL_beta4,3)`, `r round(CIU_beta4,3)`).

The estimated value of $\beta5$ is `r round(beta5_est,3)` and 95% confidence interval is (`r round(CIL_beta5,3)`, `r round(CIU_beta5,3)`).

The estimated value of $\beta6$ is `r round(beta6_est,3)` and 95% confidence interval is (`r round(CIL_beta6,3)`, `r round(CIU_beta6,3)`).

## Question 7

Interpretation:

$\beta$: The estimated value of outcome on average if the subject is always in control group is `r round(beta_est,3)` and we have 95% confidence that the ture value lies between `r round(CIL_beta,3)` and `r round(CIU_beta,3)`.

$\beta_0$: On average, the difference of estimated value of outcome between treatment group and control group in time point 0 is `r round(beta0_est,3)` and we have 95% confidence that the ture value lies between `r round(CIL_beta0,3)` and `r round(CIU_beta0,3)`.

$\beta_1$: On average, the difference of estimated value of outcome between treatment group and control group in time point 1 is `r round(beta1_est,3)` and we have 95% confidence that the ture value lies between `r round(CIL_beta1,3)` and `r round(CIU_beta1,3)`.

$\beta_2$: On average, the difference of estimated value of outcome between treatment group and control group in time point 2 is `r round(beta2_est,3)` and we have 95% confidence that the ture value lies between `r round(CIL_beta2,3)` and `r round(CIU_beta2,3)`.

$\beta_3$: On average, the difference of estimated value of outcome between treatment group and control group in time point 3 is `r round(beta3_est,3)` and we have 95% confidence that the ture value lies between `r round(CIL_beta3,3)` and `r round(CIU_beta3,3)`.

$\beta_4$: On average, the difference of estimated value of outcome between treatment group and control group in time point 4 is `r round(beta4_est,3)` and we have 95% confidence that the ture value lies between `r round(CIL_beta4,3)` and `r round(CIU_beta4,3)`.

$\beta_5$: On average, the difference of estimated value of outcome between treatment group and control group in time point 5 is `r round(beta5_est,3)` and we have 95% confidence that the ture value lies between `r round(CIL_beta5,3)` and `r round(CIU_beta5,3)`.

$\beta_6$: On average, the difference of estimated value of outcome between treatment group and control group in time point 6 is `r round(beta6_est,3)` and we have 95% confidence that the ture value lies between `r round(CIL_beta6,3)` and `r round(CIU_beta6,3)`.

## Question 8

Assumptions: 

1. Consistency; 

2. Stable Unit Treatment Value Assumption(SUTVA); 

3. Exchangeability;

4. Positivity.

## Question 9

According to estimated values of coefficients and 95% CI, we can find that $\beta, \beta_2, \beta_5, \beta_6$ are significant, which means treatments in time point 2, 5, 6 bring significant difference in outcome. 